---
title: "Clustering (example)"
author: "Zanin Pavel"
date: "May 15, 2016"
output: html_document
---

[Link to project on GitHUB](https://github.com/Filareth2015/Cluster-Analysis)     
[Link to project on RPub](http://rpubs.com/Zanin_Pavel/181052)

## About    

In this work we will show clustering the weather dataset from rattle (Williams, 2014).

## Data processing  

Loading necessary libraries:
```{r, warning=FALSE, message = FALSE, echo=TRUE, cache=TRUE}
library(rattle) # Load weather dataset. Normalise names normVarNames().
library(randomForest) # Impute missing using na.roughfix().
library(ggplot2) # Visualise the data through plots.
library(reshape2) # Reshape data for plotting.
library(fpc) #  K-means with estimating k (we will calculate average silhouette width)  
```

Loading dataset (we need only first 1000 rows):
```{r, warning=FALSE, message = FALSE, echo=TRUE, cache=TRUE}
ds <- read.csv('weatherAUS.csv', nrows = 1000)        
```

Identify the dataset:
```{r, warning=FALSE, message = FALSE, echo=TRUE, cache=TRUE}
names(ds) <- normVarNames(names(ds))
vars <- names(ds)
target <- "rain_tomorrow"
risk <- "risk_mm"
id <- c("date", "location")        
```

Ignore the IDs and the risk variable:
```{r, warning=FALSE, message = FALSE, echo=TRUE, cache=TRUE}
ignore <- union(id, if (exists("risk")) risk)        
```

Ignore variables which are completely missing:
```{r, warning=FALSE, message = FALSE, echo=TRUE, cache=TRUE}
mvc <- sapply(ds[vars], function(x) sum(is.na(x))) # Missing value count.
mvn <- names(ds)[(which(mvc == nrow(ds)))] # Missing var names.
ignore <- union(ignore, mvn)        
```

Initialise the variables:
```{r, warning=FALSE, message = FALSE, echo=TRUE, cache=TRUE}
vars <- setdiff(vars, ignore)        
```

Defining numeric variables:
```{r, warning=FALSE, message = FALSE, echo=TRUE, cache=TRUE}
inputc <- setdiff(vars, target)
inputi <- sapply(inputc, function(x) which(x == names(ds)), USE.NAMES=FALSE)

numi <- intersect(inputi, which(sapply(ds, is.numeric)))
```

Impute missing values:
```{r, warning=FALSE, message = FALSE, echo=TRUE, cache=TRUE}
if (sum(is.na(ds[vars]))) ds[vars] <- na.roughfix(ds[vars])        
```

Show size and head of dataset:
```{r, warning=FALSE, message = FALSE, echo=TRUE, cache=TRUE}
dim(ds)  

head(ds)
```

## Clusterization

The k-means algorithm is a traditional and widely used clustering algorithm.   
The algorithm begins by specifying the number of clusters we are interested in - this is the k.
Each of the k clusters is identified as the vector of the average (i.e., the mean) value of each
of the variables for observations within a cluster. A random clustering is first constructed, the
k means calculated, and then using the distance measure we gravitate each observation to its
nearest mean. The means are then recalculated and the points re-gravitate. And so on until
there is no change to the means.

A unit of distance is different for differently measure variables. For example,
one year of difference in age seems like it should be a larger difference than $1 difference in our
income. A common approach is to rescale our data by subtracting the mean and dividing by
the standard deviation. The result is that the mean for all variables is 0 and a unit of difference is one standard deviation.

### Defining the number of clusters     

For selecting k we will use Average Silhouette Width:
```{r, warning=FALSE, message = FALSE, echo=TRUE, cache=TRUE}
nk <- 1:20
model <- kmeansruns(scale(ds[numi]), krange=nk, criterion="asw")
model$bestk
```

So, in our case k = 2 is the optimum choice.

### Plot clusters

Clusters centres:
```{r, warning=FALSE, message = FALSE, echo=TRUE, cache=TRUE}
model$centers
```

Show clusters:
```{r, warning=FALSE, message = FALSE, echo=TRUE, cache=TRUE, fig.width = 10, fig.height = 6, fig.align="center"}
nclust <- 2
model <- m.kms <- kmeans(scale(ds[numi]), nclust)
dscm <- melt(model$centers)
names(dscm) <- c("Cluster", "Variable", "Value")
dscm$Cluster <- factor(dscm$Cluster)
dscm$Order <- as.vector(sapply(1:length(numi), rep, nclust))
p <- ggplot(dscm,
            aes(x=reorder(Variable, Order),
                y=Value, group=Cluster, colour=Cluster))
p <- p + coord_polar()
p <- p + geom_point()
p <- p + geom_path()
p <- p + labs(x=NULL, y=NULL)
p <- p + ylim(-1, 1)
p <- p + ggtitle("Clusters profile (variables were scaled)")
p
```